---
title: "Data Objects and Basic Visualization Activity"
output: github_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, size = "small", collapse=TRUE, comment=NA, warning = FALSE, message = FALSE)
```

## Objects used in this handout:

```{r}
babynames <- read.csv("https://raw.githubusercontent.com/deepbas/statdatasets/main/baby-names-by-state.csv")
x <- c(3,6,9,5,1)
x.mat <- cbind(x, 2*x)
x.df <- data.frame(x=x, double.x=x*2)
my.list <- list(myVec=x, myDf=x.df, myString=c("hi","bye"))
```

### Question 1: data types

- What data type is `x`? What data type is `babynames$number`? 

```{r}
typeof(x)
typeof(babynames$number)
```

- What data type is `c(x, babynames$year)`?

```{r}
typeof(c(x, babynames$year))
```

- What data type is `c(x,NA)`?  What data type is `c(x,"NA")`?

```{r}
typeof(c(x,NA))
typeof(c(x,"NA"))
```


### Question 2: Subsetting and coercion

- How can we reverse the order of entries in `x`?

```{r}
rev(x)
x[5:1]
```

- What does `which(x < 5)` equal?

```{r}
x < 5
which(x<5)
```

- What does `sum(c(TRUE,FALSE,TRUE,FALSE, FALSE))` equal?

**answer:** number of TRUE's in the vector


- What does `sum(x[c(TRUE,FALSE,FALSE,FALSE)])` equal?

**answer:** first, subset `x` to be just the first and last entries, then report their sum.

```{r}
sum(c(TRUE,FALSE,FALSE,FALSE, TRUE))
sum(x[c(TRUE,FALSE,FALSE,FALSE,TRUE)])
```


- What does `sum(x < 5)` equal?

**answer:** number of entries less than 5

```{r}
sum(x < 5)
```

- What does `sum(x[x < 5])` equal?

**answer:** sums the `x` values that are less than 5

```{r}
sum(x[x < 5])
```

- Why `dim(x.mat[1:2,1])` return `NULL` while `dim(x.mat[1:2,1:2])` returns a dimension?

**answer:** The first gives the first two rows of the first column of a matrix. Taking only one column from a matrix results in a one-dimensional vector (not a matrix) that no longer has a dimension. The second subset still has more than one row and column, so it is still a matrix and hence has a dimension.

```{r}
dim(x.mat[1:2,1])
dim(x.mat[1:2,1:2])
```



### Question 3: Lists

- Using `my.list`, show three ways to write one command that gives the 3rd entry of variable `x` in data frame `myDf`

**answer:** 

```{r}
my.list
my.list$myDf$x[3]
my.list[["myDf"]]$x[3]
my.list[[2]][3,1]
```

- What class of object does the command `my.list[3]` return?

**answer:** list
```{r}
class(my.list[3])
```


- What class of object does the command `my.list[[3]]` return?

**answer:** (atomic) character vector
```{r}
class(my.list[[3]])
```

- What class of object does the command `unlist(my.list)` return? Why are all the entries `character`s?

**answer:** (atomic) character vector 
```{r}
class(unlist(my.list))
```


### Question 4: Data Frames

Loading the `tidyverse` loads 8 packages, one of which is `ggplot2`. You can certainly load each package individually, but it has become common to simply load the `tidyverse`.

```{r}
#install.packages("tidyverse")
library(tidyverse)
babynames_MN_Amy <- babynames %>% filter(state=="MN",  name == "Amy")
```

- What is the total number of babies born in Minnesota with name `John`?

```{r}
babynames_MN_John <- babynames %>% filter(state=="MN",  name == "John")
sum(babynames_MN_John$number)
```


- In what year were highest number of babies were named Alex in Minnesota?

```{r}
babynames_MN_John$year[which.max(babynames_MN_John$number)]
```


### Question 5: Basic Plot using `ggplot2`.

```{r}
#install.packages("ggplot2")
library(ggplot2)
```

- What are the grammar of graphics needed to create a scatter-plot of the number of babies born in Minnesota named `John` vs year they were born.

- **data**: 
- **aesthetic mapping**: `y = number   `, `x =  year`
- **geometric object**: point

- Fill in the data and aesthetic mapping in the below code chunk. What is returned? What's missing?

```{r}
# Fill in the blanks
ggplot(data = babynames_MN_John , mapping = aes(x = year, y = number ))
```

- Add the appropriate geometric object to create the scatter plot. This is called adding a *layer* to a plot. Remember to always put the `+` at the end of a line, never at the start.


```{r}
ggplot(data = babynames_MN_John , mapping = aes(x = year, y = number )) +
  geom_point()
```

- Repeat the above steps with babies named `John` from Minnesota or Michigan.

```{r}
babynames_MN_MI_John <- babynames %>% filter(state=="MN" | state == "MI",  name == "John")
ggplot(data = babynames_MN_MI_John , mapping = aes(x = year, y = number )) +
  geom_point(aes(colour=state))
```

- Finally, add appropriate title and labels.

```{r fig.cap= "Count of Jonh in Minnesota Vs Michigan", tidy=FALSE}
ggplot(data = babynames_MN_MI_John , mapping = aes(x = year, y = number)) +
  geom_point(aes(colour=state))+
  xlab("Year") +
  ylab(bquote(Number~of~babies~named~.("John")))+
  ggtitle("Number of babies named John over time") +
  theme(plot.title = element_text(hjust = 0.5))    # center the plot title
```
